```{r}
library(sf)
library(terra)
library(dplyr)
library(spData)
```

# Vector attribute manipulation

```{r}
methods(class = "sf")
```

```{r}
class(world)
dim(world)
```

```{r}
world_df = st_drop_geometry(world)
class(world_df)
ncol(world_df)
```

## Attribute subsetting

```{r}
world[1:6, ]    # subset rows by position
```

```{r}
world[, 1:3]    # subset columns by positio
```

```{r}
world[1:6, 1:3] # subset rows and columns by position
```

```{r}
world[, c("name_long", "pop")] # columns by name
```

```{r, error=TRUE}
world[, c(T, T, F, F, F, F, F, T, T, F, F)] # by logical indices
# world[, 888] # an index representing a non-existent column
```

A demonstration of the utility of using logical vectors for subsetting is shown in the code chunk below. This creates a new object, small_countries, containing nations whose surface area is smaller than 10,000 km2.

```{r}
i_small = world$area_km2 < 10000
summary(i_small)
```

Note the filter applied to the rows.

```{r}
small_countries <- world[world$area_km2 < 10000, ]
small_countries
```

Alternatively

```{r}
small_countries <- subset(world, area_km2 < 10000)
small_countries
```

### Select

select() selects columns by name or position. Note "geom" included by default.

```{r}
world1 <- select(world, name_long, pop)
names(world1)
```

all columns between name_long and pop (inclusive)

```{r}
world2 <- select(world, name_long:pop)
names(world2)
```

remove specific columns with the - operator

```{r}
world3 <- select(world, -subregion, -area_km2)
names(world3)
```

Subset and rename columns at the same time with the new_name = old_name syntax

```{r}
world4 <- select(world, hr_name = name_long, population = pop, area = area_km2)
names(world4)
```

All commands return the same vector:

```         
pull(world, pop)
world$pop
world[["pop"]]
```

slice() is the row-equivalent of select()

```{r}
slice(world, 1:6)
```

filter instead of subset

```{r}
filter(world, area_km2 < 10000)
```

```{r}
filter(world, (area_km2 < 100000 & lifeExp > 82))
```

## Chaining commands

```{r}
world7 <- world |> 
  filter(continent == "Asia") |> 
  select(name_long, continent) |> 
  slice(1:5)
world7
```

## Attribute aggregation

```{r}
world_agg1 <- aggregate(pop ~ continent, FUN = sum,
                        data = world, na.rm = TRUE)
class(world_agg1)
```

[`aggregate()`](https://rspatial.github.io/terra/reference/aggregate.html) is a [generic function](https://adv-r.hadley.nz/s3.html#s3-methods) which means that it behaves differently depending on its inputs. **sf** provides the method `aggregate.sf()` which is activated automatically when `x` is an `sf` object and a `by` argument is provided

```{r}
# world_agg2 <- aggregate(world["pop"], by = list(world$continent),
#                        FUN = sum, na.rm = TRUE)
# did not work
world_agg2 <- aggregate(world["pop"], by = list(world$continent),
                        FUN = sum, na.rm = TRUE)
class(world_agg2)
dim(world_agg2)
```

```{r}
world_agg3 <- world |> 
  group_by(continent) |> 
  summarise(pop = sum(pop, na.rm = TRUE))
class(world_agg3)
```

```{r}
world_agg4 <- world |> 
  group_by(continent) |> 
  summarise(Population = sum(pop, na.rm = TRUE),
            Area = sum(area_km2),
            N = n())
class(world_agg4)
sum(world_agg4$Population)
```

```{r}
world_agg5 = world |> 
  st_drop_geometry() |>                      # drop the geometry for speed
  select(pop, continent, area_km2) |> # subset the columns of interest  
  group_by(continent) |>                     # group by continent and summarize:
  summarize(Pop = sum(pop, na.rm = TRUE), Area = sum(area_km2), N = n()) |>
  mutate(Density = round(Pop / Area)) |>     # calculate population density
  slice_max(Pop, n = 3) |>                   # keep only the top 3
  arrange(desc(N))                           # arrange in order of n. countries
world_agg5 |> print.AsIs()
```

## Attribute Joining

The most common type of attribute join on spatial data takes an `sf` object as the first argument and adds columns to it from a `data.frame` specified as the second argument.

To demonstrate joins, we will combine data on coffee production with the `world` dataset. The coffee data is in a data frame called `coffee_data` from the **spData** package (see [`?coffee_data`](https://jakubnowosad.com/spData/reference/coffee_data.html) for details). It has three columns: `name_long` names major coffee-producing nations and `coffee_production_2016` and `coffee_production_2017` contain estimated values for coffee production in units of 60-kg bags in each year. A ‘left join’, which preserves the first dataset, merges `world` with `coffee_data`.

```{r}
world_coffee <- left_join(world, coffee_data)
class(world_coffee)
```

```{r}
names(world_coffee)
```

```{r}
plot(world_coffee["coffee_production_2017"])
```

If no common variable name, specify

```{r}
coffee_renamed = rename(coffee_data, nm = name_long)
world_coffee2 = left_join(world, coffee_renamed, 
                          by = join_by(name_long == nm))
dim(world_coffee2)
```

To only get countries with coffee data

```{r}
world_coffee_inner <- inner_join(world, coffee_data)
dim(world_coffee_inner)
```

```{r}
setdiff(coffee_data$name_long, world$name_long)
```

The result shows that `Others` accounts for one row not present in the `world` dataset and that the name of the `Democratic Republic of the Congo` accounts for the other: it has been abbreviated, causing the join to miss it. The following command uses a string matching (*regex*) function from the **stringr** package to confirm what `Congo, Dem. Rep. of` should be.

```{r}
drc <- stringr::str_subset(world$name_long, "Dem*.+Congo")
drc
```

To fix this issue, we will create a new version of `coffee_data` and update the name. [`inner_join()`](https://dplyr.tidyverse.org/reference/mutate-joins.html)ing the updated data frame returns a result with all 46 coffee-producing nations.

```{r}
coffee_data$name_long[grepl("Congo,", coffee_data$name_long)] = drc
world_coffee_match <- inner_join(world, coffee_data)
dim(world_coffee_match)
```

## Creating attributes

```{r}
world_new <- world
world_new$pop_dens <- world_new$pop / world_new$area_km2
```

or...

```{r}
world_new2 <- world |> 
  mutate(pop_dens = pop / area_km2)
```

The difference between [`mutate()`](https://dplyr.tidyverse.org/reference/mutate.html) and [`transmute()`](https://dplyr.tidyverse.org/reference/transmute.html) is that the latter drops all other existing columns (except for the sticky geometry column).

### Combining columns

```{r}
world_unite <- world |> 
  tidyr::unite("con_reg", continent:region_un, 
               sep = ":", remove = TRUE)
names(world_unite)
```

### Splitting columns

```{r}
world_seperate <- world_unite |> 
  tidyr::separate(con_reg, c("continent", "region_un"), sep = ":")
names(world_seperate)
```

### Renaming

```{r}
world |> 
  rename(name = name_long) |> 
  names()
```

```{r}
new_names = c("i", "n", "c", "r", "s", "t", "a", "p", "l", "gP", "geom")
world_new_names <-  world |>
  setNames(new_names)
names(world_new_names)
```

### Dropping geometry

Don't use `-` operator

```{r}
world |> st_drop_geometry() |> class()
```

# Raster Objects

```{r}
elev = rast(nrows = 6, ncols = 6,
            xmin = -1.5, xmax = 1.5, ymin = -1.5, ymax = 1.5,
            vals = 1:36)
```

```{r}
plot(elev)
```

```{r}
grain_order <- c("clay", "silt", "sand")
grain_char <- sample(grain_order, 36, replace = TRUE)
grain_fact <- factor(grain_char, levels = grain_order)
grain <- rast(nrows = 6, ncols = 6,
              xmin = -1.5, xmax = 1.5, ymin = -1.5, ymax = 1.5,
              vals = grain_fact)
```

```{r}
plot(grain)
```

```{r}
grain2 = grain # do not overwrite the original data
levels(grain2) = data.frame(value = c(0, 1, 2), wetness = c("wet", "moist", "dry"))
levels(grain2)
```

## Raster subsetting

Raster subsetting is done with the base R operator `[`, which accepts a variety of inputs:

-   Row-column indexing
-   Cell IDs
-   Coordinates
-   Another spatial object

```{r}
# row, col
elev[1, 1]
# cell ID
elev[1]
```

modifying cells

```         
elev[1, 1] = 0
```

globbing all columns

```         
elev[]
```

Multiple cells

``` r
elev[1, c(1, 2)] <- 0
```

Replacing values of multilayered rasters can be done with a matrix with as many columns as layers and rows as replaceable cells (results not shown):

```         
two_layers = c(grain, elev) 
two_layers[1] = cbind(c(1), c(4))
two_layers[]
```

## Summarizing raster objects

```{r}
global(elev, sd) |> print.AsIs()
```

```{r}
freq(grain)
```
