```{r}
library(sf)
library(terra)
library(dplyr)
```

# Raster cropping and masking

We will use two objects to illustrate raster cropping:

-   A SpatRaster object srtm representing elevation (meters above sea level) in south-western Utah
-   A vector (sf) object zion representing Zion National Park

## Align projections

```{r}
srtm <- rast(system.file("raster/srtm.tif", package = "spDataLarge"))
zion <- read_sf(system.file("vector/zion.gpkg", package = "spDataLarge"))
```

```{r}
st_crs(srtm)
```

```{r}
st_crs(zion)
```

```{r}
zion <- st_transform(zion, st_crs(srtm))
```

```{r}
crs(zion)
```

```{r}
str(zion)
```

## Crop and mask

```{r}
srtm_cropped <- crop(srtm, zion)
srtm_masked <- mask(srtm, zion)
srtm_final <- mask(srtm_cropped, zion)
```

```{r}
srtm_final2 <- srtm |> 
  crop(zion) |> 
  mask(zion)
```

```{r}
srtm_final
```

```{r}
srtm_final2
```

Inverting the mask

```{r}
srtm_inv_masked <- mask(srtm, zion, inverse = TRUE)
```

```{r}
plot(srtm)
plot(srtm_cropped)
plot(srtm_masked)
plot(srtm_inv_masked)
```

# Raster extraction

## values at specific points

extract the values from `srtm` related to `zion_points` and create a new column in `zion_points` with these values.

```{r}
data("zion_points", package = "spDataLarge")
elevation <- terra::extract(srtm, zion_points)
zion_points <- cbind(zion_points, elevation)
```

```{r}
library(tmap)
terrain_colors <- rcartocolor::carto_pal(7, "Geyser")
tm1 <- tm_shape(srtm) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors),
            col.legend = tm_legend("Elevation (m asl)")) + 
  tm_shape(zion) +
  tm_borders(lwd = 2) + 
  tm_shape(zion_points) + 
  tm_symbols(fill = "black", size = 0.5) + 
  tm_add_legend(type = "symbols", fill = "black", size = 0.7, 
                labels = "zion_points", shape = 21)
tm1
```

## Line selectors

Example elevation profiles

```{r}
zion_transect <- cbind(c(-113.2, -112.9), c(37.45, 37.2)) |> 
  st_linestring() |> 
  st_sfc(crs = crs(srtm)) |> 
  st_sf(geometry = _)
```

```{r}
# Add id, segment the line, convert segments to points
zion_transect$id <- 1:nrow(zion_transect)
zion_transect <- zion_transect |> 
  st_segmentize(dfMaxLength = 250) |> 
  st_cast("POINT")
```

```{r}
zion_transect2 <- cbind(c(-113.2, -112.9), c(37.45, 37.2)) |> 
  st_linestring() |> 
  st_sfc(crs = crs(srtm)) |> 
  st_sf(geometry = _)
zion_transect2$id = 1:nrow(zion_transect2)
zion_transect2 = st_segmentize(zion_transect2, dfMaxLength = 250)
zion_transect2 = st_cast(zion_transect2, "POINT")
```

Derive distance between points

```{r}
zion_transect <- zion_transect |> 
  group_by(id) |> 
  mutate(dist = st_distance(geometry)[, 1])
```

Extract elevation values for each point into a column

```{r}
zion_elev <- terra::extract(srtm, zion_transect)
zion_transect <- cbind(zion_transect, zion_elev)
```

```{r}
library(ggplot2)
ggplot(zion_transect, aes(as.numeric(dist), srtm)) + 
  geom_line() +
  labs(x = "Distance (m)", y = "Elevation (m a.s.l.)") + 
  theme_bw() +
  # facet_wrap(~id) +
  theme(plot.margin = unit(c(5.5, 15.5, 5.5, 5.5), "pt"))
```

```         
library(tmap)
library(grid)
library(ggplot2)

zion_transect_line = cbind(c(-113.2, -112.9), c(37.45, 37.2)) |>
  st_linestring() |> 
  st_sfc(crs = crs(srtm)) |> 
  st_sf()

zion_transect_points = st_cast(zion_transect, "POINT")[c(1, nrow(zion_transect)), ]
zion_transect_points$name = c("start", "end")

rast_poly_line = tm_shape(srtm) +
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors),
            col.legend = tm_legend("Elevation (m)")) + 
  tm_shape(zion) +
  tm_borders(lwd = 2) + 
  tm_shape(zion_transect_line) + 
  tm_lines(col = "black", lwd = 4) + 
  tm_shape(zion_transect_points) +   
  tm_text("name", 
          text.scale = tm_scale(bg.color = "white", bg.alpha = 0.75, auto.placement = TRUE)) +
  tm_layout(legend.frame = TRUE, legend.position = c("RIGHT", "TOP"),
            legend.bg.color = "white")

plot_transect = ggplot(zion_transect, aes(as.numeric(dist), srtm)) + 
  geom_line() +
  labs(x = "Distance (m)", y = "Elevation (m a.s.l.)") + 
  theme_bw() +
  # facet_wrap(~id) +
  theme(plot.margin = unit(c(5.5, 15.5, 5.5, 5.5), "pt"))

grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2, heights = unit(c(0.25, 5), "null"))))
grid.text("A. Line extraction", vp = viewport(layout.pos.row = 1, layout.pos.col = 1))
grid.text("B. Elevation along the line", vp = viewport(layout.pos.row = 1, layout.pos.col = 2))
print(rast_poly_line, vp = viewport(layout.pos.row = 2, layout.pos.col = 1))
print(plot_transect, vp = viewport(layout.pos.row = 2, layout.pos.col = 2))
```

## Polygons

```{r}
zion_srtm_values <- terra::extract(x = srtm, y = zion)
```

```{r}
zion_srtm_values |> group_by(ID) |> 
  summarise(across(srtm, list(min = min,
                              mean = mean,
                              max = max)))
```

A similar approach works for counting occurrences of categorical raster values within polygons. This is illustrated with a land cover dataset (`nlcd`) from the **spDataLarge** package in Figure [6.4](https://r.geocompx.org/raster-vector#fig:polyextr)(B), and demonstrated in the code below:

```{r}
nlcd <- rast(system.file("raster/nlcd.tif", package = "spDataLarge"))
zion2 <- st_transform(zion, st_crs(nlcd))
zion_nlcd <- terra::extract(nlcd, zion2)
zion_nlcd |> 
  group_by(ID, levels) |> 
  count()
```

Although the **terra** package offers rapid extraction of raster values within polygons, [`extract()`](https://rspatial.github.io/terra/reference/extract.html) can still be a bottleneck when processing large polygon datasets. The **exactextractr** package offers a [significantly faster alternative](https://github.com/geocompx/geocompr/issues/813) for extracting pixel values through the `exact_extract()` function. The `exact_extract()` function also computes, by default, the fraction of each raster cell overlapped by the polygon, which is more precise (see note below for details).

```{r polyextr, echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Area used for continuous (left) and categorical (right) raster extraction.", fig.width=7.5}
rast_poly_srtm = tm_shape(srtm) + 
  tm_raster(col.scale = tm_scale_continuous(values = terrain_colors),
            col.legend = tm_legend("Elevation (m)")) + 
  tm_shape(zion) +
  tm_polygons(lwd = 2, fill_alpha = 0.3) +
  tm_title("A. Continuous data extraction") +
  tm_layout(legend.frame = TRUE, legend.position = c("LEFT", "BOTTOM"),
            legend.bg.color = "white")
rast_poly_nlcd = tm_shape(nlcd) +
  tm_raster(col.scale = tm_scale_categorical(levels.drop = TRUE),
            col.legend = tm_legend(title = "Land cover")) + 
  # tm_raster(drop.levels = TRUE, title = "Land cover", legend.show = TRUE) + 
  tm_shape(zion) +
  tm_polygons(lwd = 2, fill_alpha = 0.3)  +
  tm_title("B. Categorical data extraction") +
  tm_layout(legend.frame = TRUE, legend.position = c("LEFT", "BOTTOM"),
            legend.bg.color = "white")
tmap_arrange(rast_poly_srtm, rast_poly_nlcd, ncol = 2)
```

# Rasterization

To demonstrate rasterization in action, we will use a template raster that has the same extent and CRS as the input vector data `cycle_hire_osm_projected` (a dataset on cycle hire points in London is illustrated in Figure [6.5](https://r.geocompx.org/raster-vector#fig:vector-rasterization1)(A)) and spatial resolution of 1000 meters:

```{r}
cycle_hire_osm <- spData::cycle_hire_osm
cycle_hire_osm_projected <- st_transform(cycle_hire_osm, "EPSG:27700")
raster_template <- rast(ext(cycle_hire_osm_projected), resolution = 1000,
                        crs = crs(cycle_hire_osm_projected))
```

```{r}
rasterize(cycle_hire_osm_projected, raster_template)
```

```{r}
# default fun = "last"
ch_raster1 <- rasterize(cycle_hire_osm_projected, raster_template)
ch_raster2 <- rasterize(cycle_hire_osm_projected, raster_template,
                        fun = "length")
# capacity is a variable in the data set
ch_raster3 <- rasterize(cycle_hire_osm_projected, raster_template,
                        field = "capacity", fun = sum, na.rm = TRUE)
```

## Rasterization of lines

```{r}
california <- dplyr::filter(spData::us_states, NAME == "California")
california_borders <- st_cast(california, "MULTILINESTRING")
raster_template2 <- rast(ext(california), resolution = 0.5,
                         crs = st_crs(california)$wkt)
```

```{r}
# All cells touched by the border of the selector
california_raster1 <- rasterize(california_borders, raster_template2,
                                touches = TRUE)
# All cells whose centroids are within selector
california_raster2 <- rasterize(california, raster_template2)
```

# Spatial vectorization

## Convert centroids of raster cells into points

```{r}
elev <- rast(system.file("raster/elev.tif", package = "spData"))
elev_point <- as.points(elev) |> st_as_sf()
```

```{r new}
library(tmap)
p1 = tm_shape(elev) + 
  tm_raster(col.scale = tm_scale(n = 36)) +
  tm_title("A. Raster") +
  tm_layout(outer.margins = rep(0.01, 4), inner.margins = rep(0, 4), legend.show = FALSE)
p2 = tm_shape(elev_point) +
  tm_symbols(fill = "elev", fill.scale = tm_scale(n = 36), size = 2) +
  tm_title("B. Points") +
  tm_layout(outer.margins = rep(0.01, 4), inner.margins = rep(0.09, 4), legend.show = FALSE)
tmap_arrange(p1, p2, ncol = 2)
```

## Countour lines

```{r}
dem <- rast(system.file("raster/dem.tif", package = "spDataLarge"))
cl <- as.contour(dem) |> 
  st_as_sf()
plot(dem, axes = FALSE)
plot(cl, add = TRUE)
```
